Verifying Concurrent Search Structure Templates for C Programs using VST

### 2023-04-18 concurrency/CONCUR 2023/paper.tex
WIP
initial writing for insert of lock-coupling

To implement the template methodology for concurrent data structure implementations in C, we have implemented the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}.  Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language, as it relies on a helper function called \lstinline{findNext}. The purpose of \lstinline{findNext} is to determine the next node to be visited during the traversal. In our implementation, the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}. Alternatively, it terminates if \lstinline{findNext} fails to find a next node that includes the key value $\texttt{k}$ in the set of key values. The \lstinline{traverse} function returns a Boolean value indicating whether it has successfully found the next node during traversal. If it fails to do so, it returns false, which other functions can use to make decisions.

For instance, in a binary search tree, a node with a value of 42 would have a left child with a set of key values consisting of values less than 42 and a right child with one greater than 42. If we want to insert a new node with the target key $\texttt{k}$ into the tree, we can use the \lstinline{traverse} function to traverse the tree. When \lstinline{traverse} fails to find the next node (either to the left or right), it implies that it has reached a node with a target key $\texttt{k}$ that equals the node's value. At this point, we only have to modify the value of that node to \lstinline{value}.

@@

Figure \ref{traverse_giveup_a} depicts the give-up template algorithm as originally presented. The give-up template algorithm uses a helper function called \lstinline{inRange}, in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section. The purpose of the \lstinline{inRange} function is to determine whether the key value $\texttt{k}$ falls within the range of node $\texttt{n}$. If the key value is outside the node's range, the search is terminated, and the thread is restarted.

\wm{Describe why we need extra pieces in the struct, and how this relates to the other version.}
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
	typedef struct node_t {node *t; lock_t *lock; int min; int max;} node_t;
\end{lstlisting}

%% C implementation 
To implement the template algorithm, it is crucial to keep track of the lower and upper bounds of the keys present in the subtree rooted at the current node. Therefore, we have incorporated two new fields, \lstinline{min} and \lstinline{max}, into each node of type \lstinline{node_t} in our C implementation. These fields store the lower and upper bounds of the keys, respectively. When a thread arrives at a node in search of the key $\texttt{k}$, it checks whether $\texttt{k}$ falls within the range of that node. If $\texttt{k}$ falls within the range, the thread continues by calling \lstinline{findNext} to find the new node. If $\texttt{k}$ does not fall within the range, the thread releases the lock and returns to the root node by assigning the pointer of the root to the pointer of the current node (lines 21-22 in Figure \ref{traverse_giveup_a}).

@@

\subsubsection{Lock-coupling Template}
Proceeding with the proofs, we establish the following specification for the \lstinline{traverse} function in the lock-coupling template:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
\begin{mathpar}
	{\color{blue} {\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn->n)} 
	\ast \texttt{R(pn->n)} \ | \ \treerep\ \texttt{t} \right\rangle}}\\ 
	\texttt{traverse(pn, k)} \\ 
	\color{blue} {\langle \ \texttt{res} \ \texttt{n'}. \ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast \nodeboxrep \texttt{(pn->n')} \ast }
	\\ 
	\color{blue} {\texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range(n')})
	\   |\ \treerep\ \texttt{t} \rangle}
\end{mathpar}

@@

\begin{mathpar}
	\langle m.\ \nodeboxrep\ pn \ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k \mapsto v])\rangle

	\langle m.\ \nodeboxrep\ pn\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land m(k) = v\rangle

	\langle m.\ \nodeboxrep\ pn \ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k \mapsto \_])\rangle
\end{mathpar}

@@

>>>>

To implement the template methodology for concurrent data structure implementations in C, we have implemented the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}.  Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language, as it relies on a helper function called \lstinline{findNext}. The purpose of \lstinline{findNext} is to determine the next node to be visited during the traversal. In our implementation, the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}. Alternatively, it terminates if \lstinline{findNext} fails to find a next node that includes the key value $\texttt{k}$ in the set of key values. The \lstinline{traverse} function returns a Boolean value indicating whether it has successfully found the next node during traversal. If it fails to do so, it returns false, which other functions can use to make decisions. For instance, in a binary search tree, a node with a value of 42 would have a left child with a set of key values consisting of values less than 42 and a right child with one greater than 42. If we want to insert a new node with the target key $\texttt{k}$ into the tree, we can use the \lstinline{traverse} function to traverse the tree. When \lstinline{traverse} fails to find the next node (either to the left or right), it implies that it has reached a node with a target key $\texttt{k}$ that equals the node's value.

@@

Figure \ref{traverse_giveup_a} depicts the give-up template algorithm as originally presented. The give-up template algorithm uses a helper function called \lstinline{inRange}, in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section. 
\than{Could you write more things here to explain why we need inRange.}
The purpose of the \lstinline{inRange} function is to determine whether the key value $\texttt{k}$ falls within the range of node $\texttt{n}$. If the key value is outside the node's range, the search is terminated, and the thread is restarted. To implement the template algorithm, it is crucial to keep track of the lower and upper bounds of the keys present in the subtree rooted at the current node. Therefore, we have incorporated two new fields, \lstinline{min} and \lstinline{max}, into each node of type \lstinline{node_t} in our C implementation. These fields store the lower and upper bounds of the keys, respectively. In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys. This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes. If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.

%\wm{Describe why we need extra pieces in the struct, and how this relates to the other version.}
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
	typedef struct node_t {node *t; lock_t *lock; int min; int max;} node_t;
\end{lstlisting}

%% C implementation

When a thread arrives at a node in search of the key $\texttt{k}$, it checks whether $\texttt{k}$ falls within the range of that node. If $\texttt{k}$ falls within the range, the thread continues by calling \lstinline{findNext} to find the new node. If $\texttt{k}$ does not fall within the range, the thread releases the lock and returns to the root node by assigning the pointer of the root to the pointer of the current node (lines 21-22 in Figure \ref{traverse_giveup_a}).

@@

\subsubsection{Lock-coupling Template}
\label{traverse_proof_lock}
Proceeding with the proofs, we establish the following specification for the \lstinline{traverse} function in the lock-coupling template:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
\begin{mathpar}
	{\color{blue} {\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn->n)} 
	\ast \texttt{R(pn->n)} \ | \ \treerep\ \texttt{t} \right\rangle}}\\ 
	\texttt{traverse(pn, k)} \\ 
	\color{blue} {\langle \ \texttt{res} \ \texttt{n'}. \ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast \nodeboxrep \texttt{(pn->n')} \ast }
	\\ 
	\color{blue} {\texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range(n')})
	\   |\ \treerep\ \texttt{t} \rangle}
\end{mathpar}

@@

\begin{mathpar}
	\langle \texttt{m}.\ \nodeboxrep\ \texttt{p} \ |\ \treerep\ \texttt{m}\rangle\ \texttt{insert(p, k, v)}\ \langle \nodeboxrep\ \texttt{p}\ |\ \treerep\ (\texttt{m[k} \mapsto \texttt{v])}\rangle

	\langle \texttt{m}.\ \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{m} \rangle\ \texttt{lookup(p, k)}\ \langle \texttt{v}.\ \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{m} \land \texttt{m(k) = v}\rangle

	\langle \texttt{m}.\ \nodeboxrep\ \texttt{p} \ |\ \treerep\ \texttt{m}\rangle\ \texttt{delete(p, k)}\ \langle \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{(m[k} \mapsto \_ \texttt{])}\rangle
\end{mathpar}

@@

The \texttt{insert} method in Figure \ref{insert_lock} is the main method of lock-coupling template. It makes use of an auxiliary \texttt{traverse}  mentioned in Figure \ref{traverse_lock_b}. If \texttt{traverse} fails to find the next node (either to left or right), it implies that it has reached a node with a target key \texttt{k} that equals the node's value. At this point, we only have to modify the value of that to \texttt{value} (lines 4-5 in Figure \ref{insert_lock}). Otherwise, if \texttt{traverse} reaches a leaf, it allocates a new node with key \texttt{k} and value \texttt{v} by calling \texttt{insertOp(pn, k, v)} (line 9 in Figure \ref{insert_lock}).

Before moving on to the proof of \texttt{insert} operations for binary search tree, we present the specifications of helper functions, \texttt{findNext} is used in \texttt{traverse} function, and \texttt{insertOp} is used in \texttt{insert} function. They are all Hoare triples that operate on the local $\nodeboxrep$ predicate, meaning their proofs need not reason about concurrency.

\begin{mathpar}
	% spec of findNext
	{\color{blue}\{ \  \texttt{pn->n} \ast \texttt{n->t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} \ \}}
	\\ 
	\texttt{findNext(pn, k)} 
	\\
	{\color{blue} \{ \texttt{res n'.} \ \texttt{pn->n} \ast \texttt{n->t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)}  \ast }
	\\ 
	{\color{blue} \texttt{if res then } \texttt{(l = n'} \land \texttt{k < k')} \lor \texttt{(r = n'} \land \texttt{k > k')} 
	\texttt{ else } \texttt{(n' = n} \land \texttt{k = k')}  \}}
	\\
	\\
	%spec of insertOp
	{\color{blue} \left\{ \texttt{pn->n} \ast \texttt{n->t}  \right\}  } 
	\\
	\texttt{insertOp(pn, k, v)}
	\\
	{\color{blue} \{\texttt{t' l r lk1 lk2}. \ \ \texttt{pn->n} \ast \texttt{n->t'} \ast \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ast }
	\\
	{\color{blue} \texttt{l} \mapsto \texttt{(NULL, lk1)} \ast \texttt{r} \mapsto \texttt{(NULL, lk2)}
	\} }
\end{mathpar}

The specification of \texttt{findNext} states that a non-null \texttt{t} field points to a \texttt{node} struct containing the current node's key  \texttt{k'}, the node's value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes. When \texttt{findNext} sucessfully locates the next node \texttt{n'}, it sets \texttt{n'} as the left child node \texttt{l} if the provided key \texttt{k} is less than the current node's key \texttt{k'}. Conversely, if the given key \texttt{k} is greater than the current node's key \texttt{k'}, the next node \texttt{n'} is set as the right child node \texttt{r}. In case where \texttt{findNext} is unable to identify the next node \texttt{n'}, it retains the current node \texttt{n} as \texttt{n'},  indicating no change and that the provided key \texttt{k} is  identical to the current node's key \texttt{k'}.

The \texttt{insertOp} specification indicates that once \texttt{traverse} successfully reaches a leaf node (either the left or right node), the leaf node will have its \texttt{t} field set to \texttt{NULL}. This signifies that the leaf node currently has no key or value. At this location, a new node with the specified key \texttt{k} and value \texttt{v} should be inserted. Upon locating the appropriate leaf node for adding the new key-value pair \texttt{(k, v)}, the process generates two new empty child nodes with corresponding locks, \texttt{(NULL, lk1)} and \texttt{(NULL, lk2)}, which are pointed to by the \texttt{l} and \texttt{r} of the current node.

The atomic specification for \texttt{insert} method is
\begin{mathpar}
	{\color{blue}\langle \texttt{m}.\ \nodeboxrep\ \texttt{p} \ |\ \treerep\ \texttt{m}\rangle } \ \ \texttt{insert(p, k, v)}\ 
	{\color{blue}\langle \nodeboxrep\ \texttt{p}\ |\ \treerep\ (\texttt{m[k} \mapsto \texttt{v])}\rangle }
\end{mathpar}

Similar to the \texttt{traverse} specification, the \texttt{insert} function's specification is composed of two components: local and public pre- and postconditions. The local pre- and postconditions, containing $\nodeboxrep \ \texttt{p}$, describe the binary search tree's concrete state with the root node $\texttt{p}$, which points to the node's lock and lock invariant assertion. On the other hand, the public pre- and postconditions are represented by the predicate $\treerep \ \texttt{m}$, illustrating the binary search tree's abstract state as a collection of ghost states, with each state providing information about an individual node in the tree. Here, $\texttt{m}$ and $\texttt{m[k} \mapsto \texttt{v]}$ denote the tree's abstract states before and after the function's execution, respectively.

The proof for \texttt{insert} begins with preconditions $\nodeboxrep \ \texttt{p}$ and $\treerep \ \texttt{m}$. We set \texttt{pn->n} to point to the same memory address as \texttt{*p} (line 3 in Figure \ref{insert_lock}). After the lock for the root node is acquired (line 4 in Figure \ref{insert_lock}), the thread accesses information within the root node's lock invariant, \texttt{R(pn->n)}. As seen in Section \ref{traverse_proof_lock}, we supply the same precondition for the \texttt{traverse} function, which includes local precondition $\nodeboxrep \texttt{(pn->n)} \ast \texttt{R(pn->n)}$ and public precondition $\treerep\ \texttt{m}$. We then apply the \texttt{traverse} specification from Section \ref{traverse_proof_lock}.

In the \texttt{then} branch, when the \texttt{traverse} fails to find the next node, resulting in $\texttt{res} = \texttt{false}$, it implies that $\texttt{pn->n'->t} \neq \texttt{NULL}$. This indicates that the key to be inserted already exists in the tree. In this scenario, we update the node's value while maintaining the rest of the tree structure (line 6 in Figure \ref{insert_lock}); this acts as the linearization point for the insertion. Conversely, when the \texttt{traverse} reaches a leaf and produces $\texttt{res} = \texttt{true}$, it implies $\texttt{pn->n'->t} = \texttt{NULL}$ (line 9 in Figure \ref{insert_lock}). We then utilize the \texttt{insertOp} specification, and with enough information, we can conclude that inserting the key at this location will alter the abstract state from an unknown $\texttt{m}$ to $\texttt{m[k} \mapsto \texttt{v]}$. In both instances, we must prove that we can atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ (\texttt{m[k} \mapsto \texttt{v])}$, by demonstrating that the in-memory tree update corresponds to a local modification in the abstract tree, which inserts the new key-value pair.
